---
title: "Assignment 5 Solutions"
author: "BIOL 1007A"
date: "Winter 2023"
output:
  html_document: default
  pdf_document: default
---

## Working with matrices, lists, and data frames  

1. Assign to the variable n_dims a single random integer between 3 and 10.  
```{r}
set.seed(752)
n_dims<-sample(3:10, 1)
```

* Create a vector of consecutive integers from 1 to n_dims2. 
```{r}
vec <- seq(1:n_dims^2)
```

* Use the sample function to randomly reshuffle these values.  
```{r}
randVec <- sample(vec)
```

* create a square matrix with these elements.  
```{r}
m<-matrix(data=randVec, nrow=n_dims, ncol=n_dims)
```

* print out the matrix.
```{r, eval =F}
print(m)
```

* find a function in r to transpose the matrix.  
* print it out again and note how it has changed.
```{r, eval = F}
t(m) ##rows become columns
```

* calculate the sum and the mean of the elements in the first row and the last row.  
  
```{r eval=F}
sum(m[1,]); mean(m[1,])
sum(m[n_dims,]); mean(m[n_dims,])
```


* if have set your code up properly, you should be able to re-run it and create a matrix of different size because n_dims will change. 

This means you should not have used any integer values when setting up the matrix. Use the `n_dims` variable instead so the code can run repeatedly with different values.  

2. Create a list with the following named elements:  

* my_matrix, which is a 4 x 4 matrix filled with random uniform values  

* my_logical which is a 100-element vector of TRUE or FALSE values. Do this efficiently by setting up a vector of random values and then applying an inequality to it.  

* my_letters, which is a 26-element vector of all the lower-case letters in random order. 

```{r}
my_matrix <- matrix(data=runif(16), nrow=4, ncol=4)

my_logical <- as.logical(sample(0:1, 100, replace=T))

my_letters <- sample(letters) #or tolower(LETTERS)
```

* create a new list, which has the element[2,2] from the matrix, the second element of the logical vector, and the second element of the letters vector.  

```{r}
myList <- list(my_matrix[2,2], my_logical[2], my_letters[2])
#myList
```


* use the typeof() function to confirm the underlying data types of each component in this list  
```{r}
typeof(myList[[3]])
str(myList)
```

* combine the underlying elements from the new list into a single atomic vector with the c() function.  
* what is the data type of this vector? 
```{r}
myVec<- c(myList[[1]],myList[[2]],myList[[3]])
typeof(myVec[[3]]) # transformed to character
```
 

3. Create a data frame with following two variables (= columns) and 26 observations (= rows) below:  

* call the first variable my_unis and fill it with a random uniform value from 0 to 10  

* call the second variable my_letters and fill it with 26 capital letters in random order. 
```{r}
df<-data.frame(my_unis=runif(26, min=0, max=10),my_letters=sample(LETTERS))
```


* for the first variable, use a single line of code in R to select 4 random rows and replace the numerical values in those rows with NA. 
```{r}
df[sample(1:nrow(df), 4),1]<-NA
```

* for the first variable, write a single line of R code to identify which rows have the missing values.  
```{r eval=F}
which(!complete.cases(df))
```

* for the second variable, sort it in alphabetical order  

```{r eval=F}
df[order(df$my_letters),]
```

* calculate the column mean for the first variable.  
```{r}
mean(df$my_unis, na.rm =T)
```

* rename the second column 'my_LETTERS'
```{r}
names(df)  
names(df)[2]<- "my_LETTERS"
names(df)

```


