---
title: "Assignment 3 Potential Solutions"
author: "BIOL1007A"
date: "`r Sys.Date()`"
output: html_document
---

1. Suppose *x* = 1.1, *a* = 2.2, and *b* = 3.3. Assign each expression to the variable *z* and print the value stored in each *z*.

    a) $x^{a^b}$
    b) $(x^{a})^b$
    c) $3x^3 + 2x^2 + 1$
    
```{r}
x<-1.1
a<-2.2
b<-3.3

# a.
z<-x^(a^b)
print(z)

# b.
z<-(x^a)^b
print(z)

# c.
z<-3*x^3 + 2*x^2 + 1
print(z)
```

2. Using the `rep` and `seq` functions, create the following vectors:

    a) $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
    b) $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$
    c) $(5,4,4,3,3,3,2,2,2,2,1,1,1,1,1)$ 

```{r}
a <- c(seq(1:8), seq(from=7, to=1))
# or c(1:8,7:1) without seq
print(a)

b <- rep(1:5, c(1,2,3,4,5))
# or rep(1:5,1:5)
print(b)

c <- rep(seq(from=5, to=1), c(1,2,3,4,5))
# or rep(5:1,1:5)
print(c)
```


3. Create a vector using the code: `queue <- c("sheep", "fox", "owl", "ant")`, where `queue` represents the animals that are lined up to enter an exclusive club, with the sheep at the front of the line. Using R expressions, update `queue` as:

    a) the serpent arrives and gets in the back of the line;
    b) the sheep enters the club (so disappears from the line);
    c) the donkey arrives and talks his way to the front of the line;
    d) the serpent gets impatient and leaves;
    e) the aphid arrives and the ant invites him to cut in line (hint: check out the append function).
    f) Finally, determine the position of the aphid in the line (hint: use the which function). 
    
```{r}
queue<- c("sheep", "fox", "owl", "ant")

# the serpent arrives
queue[5]<-"serpent"
print(queue)

# the sheep enters the ark
queue<-queue[-1]
print(queue)

# the donkey arrives and talks his way to the front of the line
queue<-c("donkey",queue)
print(queue)

# the serpent gets impatient and leaves
queue<-queue[-5]
# or queue <- queue[-length(queue)]
print(queue)

# the owl gets bored and leaves
queue<-queue[-3]
# or queue <- queue[queue !="owl"]
print(queue)

# the aphid arrives and the ant invites him to cut in line
queue<-append(queue, "aphid", after=2)
# or queue <-c(queue[-length(queue)],c("aphid","ant"))
print(queue)

# Finally, determine the position of the aphid in the line.
which(queue == "aphid")
```

