---
title: "Lecture 8: Loading Data"
author: "BIOL1007A"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---


## Write a table
```{r eval=F}
write.table(x=my_data,
            file="Path/To/OutputFileName.csv",
            HEADER=TRUE,
            sep=",")
```


## Read in a table
```{r eval=F}
my_data <- read.table(file="path/to/data.csv",
                    header=TRUE,
                    sep=",",
                    comment.char="#")

```

## `saveRDS()`: useful when you're working only in R

```{r eval=F}
saveRDS(my_data, file="Path/To/FileName.RDS") # .RDS suffix is not required, but good for clarity

#This only saves a single R object as a binary, but remember, you can bundle up many things into a single list!
```

## `readRDS()`

```{r eval=F}
#Use readRDS() to restore it.
data_in <-readRDS("FileName.RDS") 
```


## 'Long' vs 'Wide' data format: 
- A wide format contains values that do not repeat in the ID column. A long format contains values that do repeat in the ID column.  
- sometimes data is collected and entered one way, but needed in a different format for analysis and plotting

### the `pivot_longer()` function in the `tidyr` package makes datasets longer by increasing the number of rows and decreasing the number of columns. 

```{r eval=F}
library(tidyverse) 

#billboard rank of songs in the year 2000
head(billboard)
billboard %>%
  pivot_longer(
    cols = starts_with("wk"), #specify which columns you want to make 'longer'
    names_to = "week", #name of new column of header names
    values_to = "rank", #name of new column of cell values
    values_drop_na = TRUE
  )
```

### `pivot_wider()` "widens" data, increasing the number of columns and decreasing the number of rows.

```{r eval=F}
head(fish_encounters) 
table(fish_encounters$seen) 
# create an occupancy matrix  
head(fish_encounters)
fish_encounters %>%
  pivot_wider(names_from = station, #which column would you like to turn into multiple columns
              values_from = seen) #which column contains the values for the new columns
# Fill in missing values with 0s
fish_encounters %>%
  pivot_wider(names_from = station, values_from = seen, values_fill = 0)

```



## Publicly available data through Dryad
#### a curated resource that makes research data discoverable, freely reusable, and citable.  
#### You can explore data sets here: https://datadryad.org/search

```{r eval=F}
# Data obtained from Veysey Powell and Babbitt 2016: An experimental test of buffer utility as a technique for managing pool-breeding amphibians https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.547rp

dryadData <- read.table("Data/veysey-babbitt_data_amphibians.csv", header=TRUE, sep=",", stringsAsFactors = TRUE)

glimpse(dryadData)
head(dryadData)
table(dryadData$species)
summary(dryadData$mean.hydro)
```


## Analyses

### Archetype Experimental Designs
- independent versus dependent variables
- discrete versus continuous variables
- continuous variables (integer and real)
- direction of cause and effect, x axis is independent
- continuous versus discrete (natural or arbitrary or statistical bins)

### Regression (dependent: continuous, independent: continuous)
- linear model of $y = a + bx$
- statistical tests for null of hypothesis of slope and/or intercept = 0
- confidence and prediction intervals of uncertainty
- goodness of fit tests for linearity


## Basic linear regression analysis in R (using 2 continuous variables)
#### Is there a relationship between mean pool hydroperiod and number of breeding frogs caught?

```{r eval=F}
# model y (response) ~ x (explanatory)
regModel <- lm(count.total.adults~mean.hydro,data=dryadData)
# model output
regModel # printed output is sparse
#View(summary(regModel)) # you can expand table to get coefficients code
head(regModel$residuals)
hist(regModel$residuals) 
# contains residuals
# 'summary' of model has elements
summary(regModel) # p=3.27e-06

summary(regModel)[["r.squared"]]
summary(regModel)$"r.squared"

summary(regModel)$coefficients

# best to examine entire matrix of coefficients:
summary(regModel)$coefficients[] #shows all
summary(regModel)$coefficients[2,1]   # slope estimate of the relationship

```


### Basic ggplot of regression model

```{r eval=F}
regPlot <- ggplot(data=dryadData,aes(x=mean.hydro,y=count.total.adults)) +
           geom_point() +
          stat_smooth(method=lm,se=0.99) # default se=0.95 
regPlot + theme_few()
# ggsave(filename="Plot1.pdf",plot=regPlot,device="pdf")
```

### Basic ANOVA in R
#### Was there a statistically signficant difference in the number of adults among years?

```{r eval=F}
# y~x 
ANOmodel <- aov(count.total.adults~factor(wetland),data=dryadData)
print(ANOmodel)
summary(ANOmodel)

dryadData %>%
  group_by(wetland) %>%
  summarise(avgHydro = mean(count.total.adults, na.rm=T), n=n())

```


### Basic ggplot of ANOVA data

```{r eval=F}
## we need to let R know that our wetland variable is categorical and not numeric
class(dryadData$wetland)
dryadData$wetland <-factor(dryadData$wetland)
class(dryadData$wetland)
ANOPlot <- ggplot(data=dryadData,aes(x=wetland,y=count.total.adults)) +
           geom_boxplot()
ANOPlot


wetlandSpCount <- ggplot(data=dryadData,aes(x=wetland,y=count.total.adults, fill=species)) +
           geom_boxplot() + scale_fill_grey()

# ggsave(filename="Plot2.pdf",plot=wetlandSpCount,device="pdf")
```


### Data frame construction for logistic regression
```{r eval=F}
xVar <- sort(rgamma(n=200,shape=5,scale=5))
yVar <- sample(rep(c(1,0),each=100),prob=seq_len(200))
lRegData <- data.frame(xVar,yVar)
```

### Logistic regression analysis in R

```{r eval=F}
lRegModel <- glm(yVar ~ xVar,
                 data=lRegData,
                 family=binomial(link=logit))
summary(lRegModel)
summary(lRegModel)$coefficients
```

### Basic ggplot of logistic regression
```{r eval=F}
lRegPlot <- ggplot(data=lRegData, aes(x=xVar,y=yVar)) +
            geom_point() +
            stat_smooth(method=glm, method.args=list(family=binomial))
print(lRegPlot)
   
```

### Data for contingency table analysis
## Counts of males and females
```{r eval=F}

countData<- dryadData %>%
  group_by(species) %>%
  summarize(Males=sum(No.males, na.rm=T), Females=sum(No.females, na.rm=T)) %>%
  select(-species) %>%
  as.matrix() 

row.names(countData)=c("SS","WF")
countData

```

### Basic contingency table analysis in R

```{r eval=F}
print(chisq.test(countData))
```

### Plotting contingency table analyses

```{r eval=F}
# some simple plots using baseR
mosaicplot(x=countData,
           col=c("goldenrod","grey"),
           shade=FALSE)

countDataLong <- countData %>%
  as_tibble() %>%
  mutate(Species=c(rownames(countData))) %>%
  pivot_longer(cols=Males:Females,
               names_to = "Sex",
               values_to = "Count")
ggplot(countDataLong, aes(x=Species, y=Count, fill=Sex)) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_manual(values=c("black", "darkslateblue"))
```


### ggplot reconstruction

```{r message=F, warning =F}
# Data obtained from Veysey Powell and Babbitt 2016: An experimental test of buffer utility as a technique for managing pool-breeding amphibians https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.547rp
library(tidyverse)
library(ggthemes)

dryadData <- read.table("Data/veysey-babbitt_data_amphibians.csv", header=TRUE, sep=",", stringsAsFactors = TRUE)

glimpse(dryadData)
head(dryadData)
table(dryadData$species)

dryadData$species<-factor(dryadData$species, labels=c("Spotted Salamander", "Wood Frog")) #creating 'labels' to use for the plot

class(dryadData$treatment)
dryadData$treatment <- factor(dryadData$treatment, 
            levels=c("Reference",
                     "100m", "30m"))


p<- ggplot(data=dryadData, 
       aes(x=interaction(wetland, treatment), 
           y=count.total.adults, fill=factor(year))) + geom_bar(position="dodge", stat="identity", color="black") +
  ylab("Number of breeding adults") +
  xlab("") +
  scale_y_continuous(breaks = c(0,100,200,300,400,500)) +
  scale_x_discrete(labels=c("30 (Ref)", "124 (Ref)", "141 (Ref)", "25 (100m)","39 (100m)","55 (100m)","129 (100m)", "7 (30m)","19 (30m)","20 (30m)","59 (30m)")) +
  facet_wrap(~species, nrow=2, strip.position="right") +
  theme_few() + scale_fill_grey() + 
  theme(panel.background = element_rect(fill = 'gray94', color = 'black'), legend.position="top",  legend.title= element_blank(), axis.title.y = element_text(size=12, face="bold", colour = "black"), strip.text.y = element_text(size = 10, face="bold", colour = "black")) + 
  guides(fill=guide_legend(nrow=1,byrow=TRUE)) 

p
```

