---
title: "Lecture 5"
author: "BIOL1007A"
date: "`r Sys.Date()`"
output: html_document
---

### Techniques for assignments and subsetting matrices and data frames
```{r,eval=FALSE}
# same principle applied to both dimensions of a matrix
m <- matrix(data=1:12,nrow=3)

print(m)
# subsetting based on elements
m[1:2,3:4]

# use blanks before or after comma to indicate full rows or columns
m[1:2, ]    
m[ ,3:4]
# use logicals for more complex subsetting
# e.g. select all columns for which the totals are > 15
# first try this logical
colSums(m) > 15
m[ , colSums(m) > 15]
# e.g. select all rows for which the row total is 22
m[rowSums(m)==22, ]
# note == for logical equal and != for logical NOT equal
m[rowSums(m)!=22, ]

# caution! simple subscripting to a vector changes the data type!
z <- m[1, ]
print(z)
str(z)
# to keep this as a matrix, must add the drop=FALSE option
z2 <- m[1, ,drop=FALSE]
print(z2)
str(z2)
# caution #2, always use both dimensions, or you will select a single matrix element
m2 <- matrix(data=runif(9),nrow=3)
print(m2)
m2[2, ]
# but now this will just pull the second element
m2[2]
# probably should specify row and column indicators
m2[2,1]
# also use logicals for assignments, not just subsetting
m2[m2>0.6] <- NA
print(m2)
# A few changes for working with data frames:
data <- iris
str(data)
# the data frame is a list of vectors, so it is set up like a matrix
data[3,2]
# you can specify just the column names
dataSub <- data[c("Species","Petal.Length")]
str(dataSub)
# or in matrix style
dataSub <- data[ ,c("Species","Petal.Length")]
str(dataSub)
# as before, with matrices, selecting only a single column changes it
# from a data frame to a vector
dataNames <- data[ ,"Species"]
str(dataNames)

# sorting by values
head(iris)
orderedIris <- iris[order(iris$Petal.Length),]
head(orderedIris)

```

### Functions
```{r eval = F}
### Functions in R
- Everything in R is a function
sum(3,2) # a "prefix" function
3 + 2 # an "operator", but it is actually a function
`+`(3,2) # the operator is an "infix" function
y <- 3
print(y)
`<-`(yy,3) # another "infix" function
print(yy)
# to see contents of a function, print it
print(read.table)
sd # shows the code
sd(c(3,2)) # call the function with parameters
# sd() # call function with default values for parameters
```

### The Anatomy Of A User-Defined Function
```
functionName <- function(argX=defaultX,argY=defaultY,argZ=defaultZ) { 
# curly bracket open marks the start of the function body
# Body of the function goes here
# Lines of R code and annotations
# May also call functions
# May also create functions
# May also create local variables
return(z)  # returns from the function a single element (z could be a list)
# curly bracket close marks the end of the function body
} 
# prints the function body
functionName 
# calls the function with default values and returns object z
functionName() 
# calls the function with user-specified values for each paramater
functionName(parX=myMatrix,parY="Order",parZ=c(0.3,1.6,2,6))
```

### Stylistic Conventions For Programming Functions
- Use prominent hash character fencing at start and at finish
- Give a header with the function name, description input, and output
- names inside functions can be short
- functions should be simple and short, no more than a screenful
- if too complex, break into multiple shorter functions
- provide default values for all function arguments
- ideally use random numbers as defaults for rapid testing



### Scoping In Functions
- global variables: visible to all parts of the code; declared in main body
- local variables: visible only within function; declared in function or passed to function through parameters
- functions "can" see variables in global environment
- global environment cannot see variables in function environment

```{r, eval=FALSE}
myFunc <- function(a=3,b=4) {
  z <- a + b
  return(z)
}

## Print vs Return: printing means showing a value in the console. Returning means giving back a value from a function. 

myFunc()
myFuncBad <-function(a=3) {
  z <- a + b
  return(z)
}

myFuncBad() # crashes because it can't find b
b <- 100

myFuncBad() # OK now because b exists as a global variable
# But it is fine to create variables locally

myFuncOK <- function(a=3) {
  bb <- 100
  z <- a + bb
  return(z)
}

myFuncOK() # no problems now
print(bb) # but this variable is still hidden from the global environment
```

### Use Multiple `return()` Statements For Different Possible Return Values
```{r, eval=FALSE}
##################################################
# FUNCTION: HardyWeinberg
# input: an allele frequency p (0,1)
# output: p and the frequencies of the 3 genotypes AA, AB, BB
#------------------------------------------------- 
HardyWeinberg<- function(p=runif(1)) {
  if (p > 1.0 | p < 0.0) {
    return("Function failure: p must be >= 0.0 and <= 1.0")
  }
  q <- 1 - p
  fAA <- p^2
  fAB <- 2*p*q
  fBB <- q^2
 vecOut <- signif(c(p=p,AA=fAA,AB=fAB,BB=fBB),digits=3)
 return(vecOut)
  }
##################################################
  HardyWeinberg()
  HardyWeinberg(p=0.25)
  HardyWeinberg(1.1) # prints error to screen
  z <- HardyWeinberg(1.1) 
  print(z) # Error message was saved to variable z!

```


### Simple regression function

```{r, eval=FALSE}
##################################################
# FUNCTION: fitLinear 
# fits simple regression line
# inputs: numeric vector of predictor (x) and response (y)
# outputs: slope and p-value
#------------------------------------------------- 
fitLinear <- function(x=runif(20),y=runif(20)) {
  myMod <- lm(y~x) # fit the model
  myOut <- c(slope=summary(myMod)$coefficients[2,1],
             pValue=summary(myMod)$coefficients[2,4])
  plot(x=x,y=y) # quick and dirty plot to check output
  return(myOut)
}
##################################################
fitLinear()
```

### Creating a more complex default value
```{r, eval=FALSE}
##################################################
# FUNCTION: fitLinear2       
# fits simple regression line
# inputs: numeric vector of predictor (x) and response (y)
# outputs: slope and p-value
#------------------------------------------------- 
fitLinear2 <- function(p=NULL) {
  if(is.null(p)) {
     p <- list(x=runif(20),y=runif(20))
  }
  myMod <- lm(p$x~p$y) # fit the model
  myOut <- c(slope=summary(myMod)$coefficients[2,1],
             pValue=summary(myMod)$coefficients[2,4])
  plot(x=p$x,y=p$y) # quick and dirty plot to check output
  return(myOut)
}
##################################################
fitLinear2()
myPars <-list(x=1:10,y=runif(10))
fitLinear2(myPars)
```



