---
title: "Entering the `tidyverse`"
author: "BIOL1007A"
date: "`r Sys.Date()`"
output: html_document
---

### Operators
> Operators are symbols that tells R to perform different operations between variables/functions/etc.  

* Assignment operator: `<-`  
* Arithmetic operators: `+` `-` `*` `/` `^` etc.  
* Logical operators: `!` `&` `|`  
* Relational operators: `==` `!=` `>` `<` `<=` `>=`  
* Miscellaneous operators: `%>%` `%in%`



### What is the `tidyverse`?
- an opinionated collection of R packages which share philosophy, grammar (how the code is structured), and data structure.  
- Some packages in the `tidyverse` are listed [here](https://www.tidyverse.org/packages/)  

```{r eval = FALSE}
## install tidyverse packages
## don't have to install them individually
## type this code in the console because you only have to install packages once
install.packages("tidyverse")

```

### Starting with the `dplyr` package

+ a new(er) package that provides a set of tools for manipulating data sets in R  
+ specifically written to be fast!   
+ has individual functions that correspond to the most common operations    
+ makes it easier for you to figure out what to do with your data  

### The core 'verbs'
1. `filter()`   
2. `arrange()`   
3. `select()`   
4. `summarize()` and `group_by()`     
5. `mutate()`   


### Start with a built-in dataset   
```{r warning=FALSE, message=FALSE, eval=F}
### whenever you want to use a package, you have to load it into your library using this code
library(tidyverse)

### Let's use a built-in data set
data(starwars)
class(starwars)

```

### What is a tibble?    
+ "a modern take on data frames"   
+ said to keep the great aspects of data frames and drops the frustrating ones (i.e. changing variable names, changing an input type)   

```{r eval=F}
glimpse(starwars) #more effective than str() in this case   
head(starwars) #looks a little different since it's a tibble (mentions dimensions and also states variable type)
```

### Cleaning up our data   
```{r eval=F}
# complete.cases is not part of dplyr, but it is very useful!
#dataClean <- starwars[complete.cases(starwars),] # removes all rows with NA values 

# to get complete cases for just certain columns 
starwarsClean <-starwars[complete.cases(starwars[,1:10]),]


# Check for NAs
is.na(starwarsClean[1,1]) #useful for only a few observations since it returns a list of True/False
anyNA(starwarsClean)
anyNA(starwars[,1:10]) # compared to our original dataset

# What does our data look like now?
glimpse(starwarsClean) # the list columns are still there but no NAs in columns 1-10
head(starwarsClean)
```


### `filter()`: Pick/subset observations by their values 
```{r eval=F}
### uses >, >=, <, <=,!=, == (not just one!) for comparisons   
### Logical operators: & | !

# filter automatically excludes NAs, have to ask for them specifically 

filter(starwarsClean, gender == "masculine" & height < 180) # you can also use commas in place of &   

filter(starwars, gender == "masculine", height <180, height > 100) # can add multiple conditions for the same variable   
```

### The `%in%` operator  

- The `%in%` 'matching' operator can be used to identify if an element (e.g., a number) belongs to a vector or data frame.   
- Using the `%in%` operator you can compare vectors of different lengths to see if elements of one vector match at least one element in another. The length of output will be equal to the length of the vector being compared (the first one). This is not possible when utilizing the == operator.

```{r eval=F}
## Simple example
# Sequences of Letters:
a <- LETTERS[1:10]
length(a)

# Second seq of letters
b <- LETTERS[4:10]
length(b)

# output depends on first vector
a %in% b
b %in% a

### using %in% to subset
filter(starwars, eye_color %in% c("blue", "brown")) # we can use %in% for a few conditions, similar to ==
```


### `arrange()`: Reorder rows   
```{r eval=F}
arrange(starwarsClean, by = height) #default is ascending order   

arrange(starwarsClean, by = desc(height)) #we can use desc() to change that

arrange(starwarsClean, height, desc(mass)) # each additional column used to break ties with the preceding column

starwars1<- arrange(starwars, height) # missing values are at the end; note we haven't been assigning anything to a variable, just printing (until now) 
tail(starwars1)

```



### `select()`: Choose variables by their names    
```{r eval=F} 
## All of these do the same thing (subset)
starwarsClean[1:10,] # using base R
select(starwarsClean, 1:10) # you can use numbers to subset
select(starwarsClean, name:species) # you can use variables names too
select(starwarsClean, -(films:starships)) # you can subset everything except particular variables

## Rearrange columns
select(starwarsClean, name, gender, species, everything()) # using the everything() helper function is useful if you have a few variables to move to the beginning
select(starwarsClean, contains("color")) ## other helpers include: ends_with, starts_with, matches (reg ex), num_range

## Renaming columns
select(starwars, haircolor = hair_color) # select can rename but only keeps specified variable; actual name of the column comes after stating new name
rename(starwarsClean,  haircolor = hair_color) # keeps all the variables
```


### `mutate()`: Create new variables with functions of existing variables   
```{r eval=F}
# Let's create a new column that is a height divided by mass
mutate(starwarsClean, ratio = height/mass) # note we use arithmetic operators 

starwars_lbs <- mutate(starwarsClean, mass_lbs=mass*2.2) # now let's convert kg to lbs
head(starwars_lbs) # mass_lbs is now the last variable of the tibble

select(starwars_lbs, 1:3, mass_lbs, everything()) # we can bring it to the beginning using select

# If we only wanted the new variable
transmute(starwarsClean, mass_lbs=mass*2.2) # just keep the new variable

transmute(starwarsClean, mass, mass_lbs=mass*2.2) # you can mention variables you want to keep in the new dataset
```


### `summarize()` and `group_by()`: Collapse many values down to a single summary 
```{r eval=F}
summarize(starwarsClean, meanHeight = mean(height)) # gives summary statistics for entire tibble

# working with NAs
summarize(starwars, meanHeight = mean(height)) # does not calculate mean if NAs are present
summarize(starwars, meanHeight = mean(height, na.rm=TRUE), TotalNumber = n()) # n() is a useful function in determining sample size
summarize(starwarsClean, meanHeight = mean(height), number = n() ) #na.rm not necessary since we are working with a clean dataset

# use group_by for maximum usefulness
starwarsGenders<- group_by(starwars, gender)
head(starwarsGenders) # now we see the groups mentioned at the top of the tibble
summarize(starwarsGenders, meanHeight = mean(height, na.rm = TRUE), number = n() ) # now we can compare height and sample size between groups
```


### What is piping? 
+ uses the forward piping operator `%>%`  
+ used to emphasize a sequence of actions   
+ it lets you pass an intermediate result onto the next function (it takes the output of one statement and makes it the input of the next statement)
+ avoid when you need to manipulate more than one object at a time or there are meaningful intermediate objects   
+ formatting: should always have a space before it and usually followed by a new line (usually automatic indent)

```{r eval=F}
starwarsClean %>%
    group_by(gender) %>%
    summarize(meanHeight = mean(height, na.rm=T), number=n()) 
# so much cleaner! 
```

```{r eval=F}
## case_when() is useful for multiple ifelse statements

starwarsClean %>%
  mutate(sp = case_when(species == "Human" ~ "Human", TRUE ~ "Non-Human")) %>%
  select(name, sp, everything())

starwarsClean %>%
  group_by(films) %>%
  mutate(
    sp = case_when(species == "Human" ~ "Human", TRUE ~ "Non-Human"),
    status = case_when(str_detect(films, "A New Hope") ~ "OG", TRUE ~ "Later")) %>%
  select(name,sp, status, everything()) %>%
  arrange(status) %>%
  {starwarsOGstatus <<- .}

tail(starwarsOGstatus)
```

```{r eval=F}
## long to wide format and vice-versa
glimpse(starwarsClean)

# make data set wider
wideSW <- starwarsClean %>% 
  select(name,sex, height) %>%
  pivot_wider(names_from = sex, values_from = height, values_fill = NA)
wideSW

## make data set longer
glimpse(wideSW)
wideSW %>%
  pivot_longer(cols= male:female,
               names_to = "sex",
               values_to= "height",
               values_drop_na = T)

# one last example
starwars %>%
  select(name, homeworld) %>%
  group_by(homeworld) %>%
  mutate(rn = row_number()) %>%
  ungroup() %>%
  pivot_wider(names_from = homeworld, values_from = name) %>%
  select(-rn)

  
```
