---
title: "Simple Data Analyses and More Complex Control Structures"
author: "BIOL1007A"
date: "`r Sys.Date()`"
output: html_document
---

## Load in data set

```{r eval=F}
# Data obtained from Veysey Powell and Babbitt 2016: An experimental test of buffer utility as a technique for managing pool-breeding amphibians https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.547rp
library(tidyverse)
dryadData <- read.table("Data/veysey-babbitt_data_amphibians.csv", header=TRUE, sep=",")

glimpse(dryadData)
```


## Analyses

### Archetype Experimental Designs
- independent versus dependent variables
- discrete versus continuous variables
- continuous variables (integer and real)
- direction of cause and effect, x axis is independent
- continuous versus discrete (natural or arbitrary or statistical bins)

### Regression (dependent: continuous, independent: continuous)
- linear model of $y = a + bx$
- statistical tests for null of hypothesis of slope and/or intercept = 0
- confidence and prediction intervals of uncertainty
- goodness of fit tests for linearity


## Basic linear regression analysis in R (using 2 continuous variables)
#### Is there a relationship between mean pool hydroperiod and number of breeding frogs caught?

```{r eval=F}
# model y (response) ~ x (explanatory)
regModel <- lm(count.total.adults~mean.hydro,data=dryadData)
# model output
regModel # printed output is sparse
#View(summary(regModel)) # you can expand table to get coefficients code
head(regModel$residuals)
hist(regModel$residuals) 
# contains residuals
# 'summary' of model has elements
summary(regModel) # p=3.27e-06

summary(regModel)[["r.squared"]]
summary(regModel)$"r.squared"

summary(regModel)$coefficients

# best to examine entire matrix of coefficients:
summary(regModel)$coefficients[] #shows all
summary(regModel)$coefficients[2,1]   # slope estimate of the relationship

```


### Basic ggplot of regression model

```{r eval=F}
library(ggthemes)
regPlot <- ggplot(data=dryadData,aes(x=mean.hydro,y=count.total.adults)) +
           geom_point() +
          stat_smooth(method=lm,se=0.99) # default se=0.95 
regPlot + theme_few()
# ggsave(filename="Plot1.pdf",plot=regPlot,device="pdf")
```

### Basic ANOVA in R
#### Was there a statistically signficant difference in the number of adults among years?

```{r eval=F}
# y~x 
ANOmodel <- aov(count.total.adults~factor(wetland),data=dryadData)
print(ANOmodel)
summary(ANOmodel)

dryadData %>%
  group_by(wetland) %>%
  summarise(avgHydro = mean(count.total.adults, na.rm=T), n=n())

```


### Basic ggplot of ANOVA data

```{r eval=F}
## we need to let R know that our wetland variable is categorical and not numeric
class(dryadData$wetland)
dryadData$wetland <-factor(dryadData$wetland)
class(dryadData$wetland)
ANOPlot <- ggplot(data=dryadData,aes(x=wetland,y=count.total.adults)) +
           geom_boxplot()
ANOPlot

# by species
wetlandSpCount <- ggplot(data=dryadData,aes(x=wetland,y=count.total.adults, fill=species)) +
           geom_boxplot() + scale_fill_grey()
wetlandSpCount
# ggsave(filename="Plot2.pdf",plot=wetlandSpCount,device="pdf")
```


### Data frame construction for logistic regression
```{r eval=F}
## Gamma Distribution: a continuous probability distribution that is widely used in different fields of science to model continuous variables that are always positive and have skewed distributions.

xVar <- sort(rgamma(n=200,shape=5,scale=5))
yVar <- sample(rep(c(1,0),each=100),prob=seq_len(200))
lRegData <- data.frame(xVar,yVar)
```

### Logistic regression analysis in R

```{r eval=F}
lRegModel <- glm(yVar ~ xVar,
                 data=lRegData,
                 family=binomial(link=logit))
summary(lRegModel)
summary(lRegModel)$coefficients
```

### Basic ggplot of logistic regression
```{r eval=F}
lRegPlot <- ggplot(data=lRegData, aes(x=xVar,y=yVar)) +
            geom_point() +
            stat_smooth(method=glm, method.args=list(family=binomial))
print(lRegPlot)
   
```

### Data for contingency table analysis
## Counts of males and females between species
```{r eval=F}

countData<- dryadData %>%
  group_by(species) %>%
  summarize(Males=sum(No.males, na.rm=T), Females=sum(No.females, na.rm=T)) %>%
  select(-species) %>%
  as.matrix() 

row.names(countData)=c("SS","WF")
countData

```

### Basic contingency table analysis in R

```{r eval=F}
print(chisq.test(countData))
```

### Plotting contingency table analyses

```{r eval=F}
# some simple plots using baseR
mosaicplot(x=countData,
           col=c("goldenrod","grey"),
           shade=FALSE)

countDataLong <- countData %>%
  as_tibble() %>%
  mutate(Species=c(rownames(countData))) %>%
  pivot_longer(cols=Males:Females,
               names_to = "Sex",
               values_to = "Count")
ggplot(countDataLong, aes(x=Species, y=Count, fill=Sex)) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_manual(values=c("black", "darkslateblue"))
```

## Touching on some more advanced control structures

### `if` statements

```
if (condition) {expression1}
if (condition) {expression1} else {expression2}
if (condition1) {expression1} else
if (condition2) {expression2} else
```
* note that final unspecified `else` captures rest of the (unspecified) conditions
* `else` statement must appear on the same line as the expression
* typically enclose multiple statements in brackets `{}` for compound expression

```{r, eval=FALSE}
z <- signif(runif(1),digits=2)
print(z)
z > 0.5

# works with or without {} after if
if (z > 0.8) {cat(z,"is a large number","\n")} else 
if (z < 0.2) {cat(z,"is a small number","\n")} else
   {cat(z,"is a number of typical size","\n")
    cat("z^2 =",z^2,"\n")}

# if statement requires a single logical value. With a vector,
# it will throw an error.
z <- 1:10
# this does not do anything
if (z > 7) print(z)

# use subsetting!
print(z[z < 7])
   
```

# `ifelse` to fill vectors
```
ifelse(test,yes,no)
test is an object that can be coerced to a logical yes/no
yes return values for true elements of test
no return vales for false elements of test
```
Suppose we have an insect population in which each female lays, on average, 10.2 eggs, following a Poisson distribution with $\lambda = 10.2$. However, there is a 35% chance of parasitism, in which case no eggs are laid. Here is the distribution of eggs laid for 1000 individuals:

```{r, eval=FALSE}
# Poisson: a statistical distribution showing the likely number of times that an event will occur within a specified period of time

tester <- runif(1000) # start with random uniform elements
eggs <- ifelse(tester>0.35,rpois(n=1000,lambda=10.2),0)
hist(eggs)


```

Suppose we have a vector of p values (say from a simulation), and we want to create a vector to highlight the significant ones for plotting purposes.

```{r}
pVals <- runif(1000)
z <- ifelse(pVals<=0.025,"lowerTail","nonSig")
z[pVals>=0.975] <- "upperTail"
table(z)
```


#### `for loops`

### `for` loops
* The workhorse function for doing repetitive tasks
* Universal in all computer languages
* Controversial in R
    - often not necessary (use vectorized operations!)
    - very slow with binding operations (c,rbind,cbind,list) 
    - many operations can be handled by special family of `apply` functions
    
#### Anatomy of a `for` loop

```
for (var in seq) { # start of for loop
# body of for loop 
} # end of for loop
var is a counter variable that will hold the current value of the loop
seq is an integer vector (or a vector of character strings) that defines the starting and ending values of the loop
```

It is traditional in the statistics literature to use variables `i,j,k` to indicate counters. e.g.

$$A = \sum_{i=1}^S n_i$$

There are many ways to indicate the sequence, and the length of our loops will almost always be related to the size of a vector that we are working with. 

We will only illustrate one way, because it is universal and applies to all for loops. 

```{r}
for (i in 1:5) {
  cat("stuck in a loop",i,"\n")
  cat(3 + 2,"\n")
  cat(runif(1),"\n")
}
print(i)
```


  Instead, we want to use a counter variable that maps to the position of each element
 
```{r}
my_dogs <- c("chow","akita","malamute","husky","samoyed")
for (i in 1:length(my_dogs)){
  cat("i =",i,"my_dogs[i] =" ,my_dogs[i],"\n")
}
```

### Using double for loops

```{r, eval=TRUE}
m <- matrix(round(runif(20),digits=2),nrow=5)
# loop over rows
for (i in 1:nrow(m)) { # could use for (i in seq_len(nrow(m)))
  m[i,] <- m[i,] + i
} 
print(m)

# Loop over columns
m <- matrix(round(runif(20),digits=2),nrow=5)
for (j in 1:ncol(m)) {
  m[,j] <- m[,j] + j
}
print(m)

# Loop over rows and columns
m <- matrix(round(runif(20),digits=2),nrow=5)
for (i in 1:nrow(m)) {
  for (j in 1:ncol(m)) {
    m[i,j] <- m[i,j] + i + j
  } # end of column j loop
} # end or row i loop
 print(m) 
```


